---
title: "AEF assignment 1#"
author: "Martin, Fred and Oliver"
date: "6/3/2022"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=FALSE, warning=FALSE, message=FALSE}
# Loading packages:
library(RSQLite)
library(dbplyr)
library(dplyr)
library(tidyverse)
library(lubridate)
library(sandwich)
library(lmtest)
library(scales)
library(slider)
library(furrr)

```

### 1. Read in the monthly CRSP date (crsp_monthly) and the monthly market excess returns (factors_ff_monthly) from the tidy_finance.sqlite database. Very briefly state the definition of the variables permno, ret_excess and mktcap. Provide suitable summary statistics of the cross-section of monthly excess returns and market capitalization in the CRSP sample.




```{r}
# ******************************************
# 4.1 Data preparation
# ******************************************

#Connecting to the database
tidy_finance <- dbConnect(SQLite(), "data/tidy_finance.sqlite",
                          extended_types = TRUE
      )

#Reading in crsp_monthly dataset
crsp_monthly <- tbl(tidy_finance, "crsp_monthly") %>%
  collect() %>%
  mutate(month = as.Date(as.POSIXct.Date(month)))

#Reading in factors_ff_monthly dataset
factors_ff_monthly <- tbl(tidy_finance, "factors_ff_monthly") %>%
  collect() %>%
  mutate(month = as.Date(as.POSIXct.Date(month)))

#Merging crsp_monthly and factors_ff_monthly through a left-join, merging by the common "month" column 
crsp_monthly <- crsp_monthly %>%
  left_join(factors_ff_monthly, by = "month") %>%
  select(permno, month, ret_excess, mkt_excess, mktcap,mktcap_lag) #selecting these columns from our joint set


dbListTables(tidy_finance) # check the database
```
*Very briefly state the definition of the variables permno, ret_excess and mktcap*. 
Definition of PERMNO, as per {PERMNO}: "PERMNO is a unique permanent security identification number assigned by CRSP to each security". Thus PERMNO is a unique company identifier. 

Definition of excess returns, as per {exret}: Excess return for a stock is the difference between the stock return and the return on the risk-free security over the same period. For such risk-free securities, it is common to use a government bond such as a 10-year US Treasury bond.\\ 

Definition of mktcap, as per {mktcap}: The variable 'mktcap' is defined as the market capitalization, which refers to the total dollar market value of a company's outstanding shares of stock. Market capitalization is commonly referred to as "market cap," and is calculated by multiplying the total number of a company's outstanding shares by the current market price of one share.

*Provide suitable summary statistics of the cross-section of monthly excess returns and market capitalization in the CRSP sample*
```{r}
#Creating summary statistics for the excess return
ret = crsp_monthly %>% 
  group_by(month) %>%
  summarise(across(
    ret_excess, #listing which statistics we want included
    list(daily_mean = mean,
         daily_sd = sd,
         daily_min = min,
         daily_max = max
         ) 
  )) %>% summarise(across(-month,mean)) %>%
  rename(mean=ret_excess_daily_mean, sd=ret_excess_daily_sd, min=ret_excess_daily_min, max=ret_excess_daily_max) %>%
  pivot_longer(
    everything()
  ) %>% 
  rename(ret_excess=value) %>%
  mutate(ret_excess = format(ret_excess, scientific = F, digits=2))

#Creating summary statistics for the market cap
mkt = crsp_monthly %>% 
  group_by(month) %>%
  summarise(across(
    mktcap, #listing which statistics we want included
    list(daily_mean = mean,
         daily_sd = sd,
         daily_min = min,
         daily_max = max
         ) 
  )) %>% summarise(across(-month,mean)) %>%
  rename(mean=mktcap_daily_mean, sd=mktcap_daily_sd, min=mktcap_daily_min, max=mktcap_daily_max) %>%
   pivot_longer(
    everything()
  )  %>% 
  rename(mktcap=value,n=name) %>%
  mutate(mktcap = format(mktcap, scientific = F, digits=2))

#Merging the summary statistics so that they are presented for both excess returns and market cap
ret %>% 
  cbind(mkt) %>%
  select(-n)


```
Summary statistics for ret_excess and mktcap shows that the mean return per month is 0.79% although the median is -0.0004%. An important distinction between the variables is that ret_excess is in relative numbers, while mktcap of course is absolute, since the return is a relative change and market cap is a true number. 

#### 2. Generate a new column ret_excess_lag which contains the previous month excess return for the same firm for each row in the sample. Briefly state why crsp_monthly %>% group_by(permno) %>% mutate(ret_excess_lag = lag(ret_excess)) does not provide the correct result if crsp_monthly contains implicitly missing values (i.e. permno-month combinations that are simply not present in the data). Report suitable summary statistics or visualizations that show if returns rt and rt−1 exhibit autocorrelation that is significantly different from zero. What do your results imply? Is there short-term momentum?



### 3. Generate a new column mktcap_lag_12 which contains the firms’ market capitalization 12-month prior to the measurement date.

Lagging a variable can be a frustrating task, since the days of a month will differ. The below function will always return a date in the nth month after Date. This allows for 1) not exceeding the end of month and 2) handles NA's really well. 

First we create a variable by lagging the month column without touching the others. Next, we merge our sorting variable with the return data. We use the 12-month lagged marketcap as a sorting variable to ensure that the sorts rely only on information available when we create the portfolios.
```{r}
#Creating the 12-month lagged mktcap variable by mutating the month col
mktcap_lag_12 <- crsp_monthly %>%
  mutate(month = month %m+% months(12)) %>%
  select(permno, month, mktcap_lag_12 = mktcap) %>%
  drop_na()

#Adding lagged mktcap to existing crsp_monthly dataset through a left join 
 data_for_sorting <- crsp_monthly %>%
  left_join(mktcap_lag_12, c("permno","month")) %>%
   drop_na()
```

### Compute the momentum of stock i as the relative change in market capitalization, represented in percentage terms (1.00 is a 1% return) during the period covering months t−12 until t−1. Specifically, if $mc_{i,t}$ denotes the market capitalization of stock i in month t, compute momentum as  
$$Mom_{i,t} = 100\frac{(mc_{i,t-1} −mc_{i,t-12})}{mc_{i,t-12}}$$
```{r}
#Adding momentum column based on above formula in our dataset with lagged mktcap
data_for_sorting <-  data_for_sorting %>%
  mutate(momentum =  100*(mktcap_lag-mktcap_lag_12)/mktcap_lag_12)
```
### Briefly discuss the difference between computing momentum as the relative change in prices or as the relative change in market capitalization.*
Since market cap is basically the stock price times volume, there is rarely a large distinction between the two. But the change in market cap will always reflect a change in the public value of a company whereas a change in stock price might be due to stock split or buybacks, which would have significant implications for the stock price but not the actual market capitalization (unless prices are given in adjusted form, e.g. as per yahoo! finance). 

Create summary statistics for $Mom_{i,t}$  as follows: Each month, compute the mean, standard deviation,
minimum, fifth percentile, 25th percentile, median, 75th percentile, 95th percentile, and maximum values
of the cross-sectional distribution of $Mom_{i,t}$. 

```{r}
#Creating summary statistics for each month, i.e. 700+ rows of summary statistics
mom_stat = data_for_sorting %>% 
  group_by(month) %>%
  summarise(across(
    momentum, #For the momentum column, summarise the following elements in list()
    list(Mean = mean,SD = sd, Min = min,
         Q5 = ~quantile(., 0.05),
         Q25 = ~quantile(., 0.25),
         Median = median,
         Q75 = ~quantile(., 0.75),
         Q95 = ~quantile(., 0.95),
         Max = max
         ),
    .names = "{.fn}")) #Hvad gør denne præcis ved navnene?

```

*Then, report the time-series means for each cross-sectional
value. What is the mean value of $Mom_{i,t}$ in the average month? What is the cross-sectional standard deviation? 

Does momentum exhibit a positive correlation with $log(mc_{i,t}$?*

```{r}
#Summarise the previously created montly summary statistics into one row by taking the mean in each column
mom_stat %>% 
  summarise(across(-month,mean)) 

#Plotting correlation matrix for momentum and log_mktcap 
data_for_sorting %>%
  mutate(log_mktcap = log(mktcap)) %>%
  select(momentum,log_mktcap) %>%
  cor()
  
```
Yes, there is a positive correlation coefficient of 0.067 between the momentum value and log of market capitalization. 

### 4. Next, examine the relation between momentum and future stock returns. 
** For that purpose, perform
univariate portfolio sorts in the following way: **

* Compute monthly portfolio breakpoints. Each month, use $Mom_{i,t}$ as a sorting variable. The portfolio
breakpoints are the deciles of $Mom_{i,t}$ calculated using all stocks i in the sample during month t.

We need to create a function that is able to sort stocks into a number of portfolios. We use quantile() to compute breakpoints for n_portfolios. Then, we assign portfolios to stocks using the findInterval() function. The output of the following function is a new column that contains the number of the portfolio to which a stock belongs.

First, we create a function that is able to sort stocks into a number of portfolios based on their momentum score. The output one gets when running the function with a dataframe is a new column that contains the number of the portfolio in which the stock belongs. 

```{r, echo=FALSE}
assign_portfolio <- function(data, var, n_portfolios) {
  breakpoints <- data %>%
    summarize(breakpoint = quantile({{ var }},
                                    probs = seq(0, 1, length.out = n_portfolios + 1),
                                    na.rm = TRUE
    )) %>%
    pull(breakpoint) %>%
    as.numeric()
  
  data %>%
    mutate(portfolio = findInterval({{ var }},
                                    breakpoints,
                                    all.inside = TRUE
    )) %>%
    pull(portfolio)
}
```

```{r}
#We can use the above function to sort stocks into ten portfolios each month using lagged betas and compute value-weighted returns for each portfolio. Note that we transform the portfolio column to a factor variable because it provides more convenience for the figure construction below.

  momentum_portfolios = data_for_sorting %>%
  group_by(month) %>%
  mutate(
    portfolio = assign_portfolio(
      data = cur_data(),
      var = momentum,
      n_portfolios = 10
    ),
    portfolio = as.factor(portfolio)
  )

momentum_portfolios%>%
  head()
  
  

```
* To help understand the characteristics of the stocks comprising each of the $Mom_{i,t}$ sorted portfolios,
present the equal-weighted average values of $Mom_{i,t}$ and $mc_{i,t}$ for each of the 10 portfolios.

```{r}

momentum_portfolios %>% 
  group_by(portfolio) %>%
  summarise(
    mom_mean = mean(momentum),
    mc_mean = mean(mktcap)
    ) 
```

* Compute value-weighted monthly excess returns for the decile portfolios.
* Report the average excess return, the CAPM alpha and the market beta for the 10 momentum sorted portfolios. 

```{r}
vw_momentum_portfolios = momentum_portfolios %>%
  group_by(portfolio, month) %>%
  summarize(excess_ret = weighted.mean(ret_excess, mktcap_lag), .groups = "drop") # The value weighted return


mom_portfolios_summary <- vw_momentum_portfolios %>%
  left_join(factors_ff_monthly, by = "month") %>%
  group_by(portfolio) %>%
  summarise(
    alpha = as.numeric(lm(excess_ret ~ 1 + mkt_excess)$coefficients[1]),
    beta = as.numeric(lm(excess_ret ~ 1 + mkt_excess)$coefficients[2]),
    excess_ret = mean(excess_ret)
  )
mom_portfolios_summary
```

```{r}

mom_portfolios_summary %>%
  ggplot(aes(x = portfolio, y = alpha, fill = portfolio)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Alphas of momentum-sorted portfolios",
    x = "Portfolio",
    y = "CAPM alpha",
    fill = "Portfolio"
  ) +
  scale_y_continuous(labels = percent) +
  theme(legend.position = "None")
```

The plot shows that portfolios 1 and 10 has the highest alpha and decreasing going towards the center. Thus alpha is lowest for median momentum stocks. 


**Finally, analyse the momentum strategy: a portfolio that goes long past winners (the highest decile) and short past
losers (the lowest decile).**
```{r}
strategy_portfolio = momentum_portfolios %>% 
  mutate(portfolio = as.numeric(portfolio)) %>%
  group_by(month) %>%
  mutate(breakpoint_low = 1,
         breakpoint_high = 10,
         portfolio = case_when(portfolio <= breakpoint_low ~ "low",
                               portfolio >= breakpoint_high ~ "high")) %>%
  group_by(month,portfolio) %>%
  summarise(excess_ret = weighted.mean(ret_excess, mktcap_lag)) # Value weighted return by high or low grouped momentum values. 

strategy_portfolio %>% 
  pivot_wider(names_from = portfolio, values_from = excess_ret) %>%
  mutate(high_low = high - low) %>% 
  left_join(factors_ff_monthly, by = "month") %>%
  lm(high_low ~ mkt_excess, data = .) %>% broom::tidy()
```

* What do you conclude about the performance of the momentum strategy?
Is the strategy market neutral (i.e., does it exhibit a market beta which is zero)? Does the strategy deliver abnormal excess returns (i.e., a positive alpha)?

The test statistics shows that is it possible to generate (very) small alpha by having a short position in low momentum stocks and a long position in high momentum stocks. The alpha of this regression is 0,14%, which may be considered as insignificant. The insignificant alpha can be implied by the graph above, which shows that alpha is highest within the lowest- and highest momentum portfolios. A strategy in this case would be to go long these portfolios and short the rest.
The strategy is not market neutral as the beta is -0.1. 

#### One of the main conclusions in the momentum literature is that the timing of the measurement of momentum plays a substantial role in the nature of the relation between momentum and future stock returns.
To assess this in more detail and to examine whether momentum has the ability to predict returns further in the future than the next month, repeat the univariate portfolio sorts but this time using the k-month-ahead excess return as the outcome variable for values of k ∈{1, 3, 6, 12}.

**For which time-horizon does the momentum deliver the highest risk-adjusted performance?**
The momentum strategy delivers the highest risk-adjusted excess return with a 1-month ahead horizon. 

```{r}

portfolio = function(data = data) {
  momentum_portfolios_inc_lagged <- momentum_portfolios %>%
  left_join(data, c("permno","month")) %>%
   drop_na()
 
strategy_portfolio = momentum_portfolios_inc_lagged %>% 
  mutate(portfolio = as.numeric(portfolio)) %>%
  group_by(month) %>%
  mutate(breakpoint_low = 1,
         breakpoint_high = 10,
         portfolio = case_when(portfolio <= breakpoint_low ~ "low",
                               portfolio >= breakpoint_high ~ "high")) %>%
  group_by(month,portfolio) %>%
  drop_na() %>%
  summarise(ret_excess_k_month_ahead = weighted.mean(ret_excess_k_month_ahead, mktcap_lag)) # Value weighted return by high or low grouped momentum values. 

estimate = strategy_portfolio %>% 
  pivot_wider(names_from = portfolio, values_from = ret_excess_k_month_ahead) %>%
  mutate(high_low = high - low) %>% 
  left_join(factors_ff_monthly, by = "month") %>%
  lm(high_low ~ mkt_excess, data = .) %>% broom::tidy()
}

```

```{r}

k = c(1,3,6,12)
data = matrix(nrow=length(k),ncol = 2)
colnames(data) <- c("k ahead", "alpha")


for (i in seq_along(k) ) {

 lag_excess_ret <- momentum_portfolios %>%
  mutate(month = month %m+% months(k[[i]])) %>%
  select(permno, month, ret_excess_k_month_ahead = ret_excess)
 
 estimate = portfolio(lag_excess_ret)
 


data[i,"alpha"] = estimate$estimate[1]
data[i,"k ahead"] = k[[i]]

}
```

```{r}
test = data %>% as.data.frame() %>% rename(k = 'k ahead')
test %>%
  ggplot(aes(x=k, y=alpha, fill=k)) + 
  geom_bar(stat = "identity") +
  labs(
    title = "Alphas of momentum-sorted portfolios k month-ahead",
    x = "k months ahead",
    y = "CAPM alpha",
    fill = "Portfolio"
  ) +
  scale_y_continuous(labels = percent) +
  theme(legend.position = "None")
``` 

### 6. Most academic studies ignore real world costs and other forms of slippage when examining factors, which is likely a larger issue for momentum due to its higher turnover. The concern is that momentum is so costly to trade that its return premium is diminished in the real world. 

#### First, based on your findings in 1. and 3., briefly explain why you think a momentum strategy could be costlier to trade than, say, a strategy based on size sorts.
When a given portfolio selection is based on i.e., past year market cap momentum, then the composition of the portfolio would change frequently as this is based purely on the performance of the stocks. Basically, since the analysis is made purely on activity from the past 12 months the portfolio composotion will change often. Compared to a beta selection strategy where the analysis is made from i.e., 20 or 30 year historic stock price data, then the estimate would change much less frequently. This dramatically increases the need for trading activity and thus transaction costs of the momentum strategy. And for the momentum strategy to be efficient, there will have to be frequent turnover, otherwise the strategy might not be able to catch it if a stock loses momentum after it's traded.  



#### Then, propose an alternative strategy that may be able to effectively capture the momentum premium but which at the same time delivers lower trading costs, e.g. because the turnover is smaller, the strategy takes less extreme (short) positions or is tilted towards stocks that are cheaper to trade. 

Describe the strategy and implement it within the CRSP sample. Report the resulting Sharpe-ratio and average turnover and compare to the baseline momentum strategy from exercise 4.
