---
title: "Assignment3"
author: "Martin, Oliver and Fred"
date: "6/5/2022"
output: html_document
---

```{r setup, include=FALSE, }
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Loading packages:
library(RSQLite)
library(tidyverse)
library(tidymodels) 
library(furrr) 
library(glmnet)
library(broom)
library(timetk)
library(scales)
library(lubridate)
library(scales)
library(frenchdata)
library(caret)
library(ggplot2)
library(ggpubr)
library(kableExtra)
library(mltools)
library(data.table)
library(vtable)
library(keras)
```


```{r}
# Select all stocks from the CRSP universe
 tidy_finance <- dbConnect(SQLite(), "data/tidy_finance.sqlite", 
                           extended_types = TRUE)

# Reading in crsp_monthly dataset
crsp_monthly <- tbl(tidy_finance, "crsp_monthly") %>%
  select(permno,month,ret_excess,industry) %>%
  collect() %>%
  mutate(month = as.Date(as.POSIXct.Date(month)))

remove(tidy_finance)
```
To keep the analysis simple we work with a balanced panel and exclude tickers, which is not traded for the full period. 
```{r}

# Clean data by removing stocks, which is not traded on every single trading day. 
data <- crsp_monthly %>%
  group_by(permno) %>%
  mutate(n=n()) %>%
  ungroup() %>%
  filter(n == max(n)) %>%
  select(-n)

remove(crsp_monthly)
```

```{r, fig.align = 'center'}
# Provide a brief summary of the sample:

table = data %>% 
  summarise(
        Variable = "Ret_excess",
        "N stocks"  = n_distinct(month),
        Mean = mean(ret_excess),
        SD = sd(ret_excess),
        Min = min(ret_excess),
        Max = max(ret_excess),
        Start = min(month),
        End = max(month)
         ) %>%
 mutate(Mean = format(Mean, digits = 2), SD = format(SD, digits = 2),
        Min = format(Min, digits = 3), Max = format(Max, digits = 3))

ggtable = ggtexttable(table, rows = NULL,
                       theme = ttheme("classic")) %>%
  tab_add_title(text = "Summary statistics")

plot = data %>% 
  select(permno, industry) %>%
  distinct() %>%
  group_by(industry) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = industry, y = n)) +
  ggtitle("Number of stocks per industry") +
  geom_bar(stat="identity", width=.3, fill="tomato3") +
  labs(x="Industry class",
         y="Number of stocks") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1))

# We can arrange the two plots to show them together in one graph. 
ggarrange(plot, ggtable, nrow = 2,
                   heights = c(1, 0.5)) 



remove(table,plot,ggtable, lineplot, figure)

```

### 2.1 Derive a closed-form solution for the mean-variance efficient portfolio ω∗t+1 based on the transaction cost specification above. Discuss if the assumption of transaction costs proportional to volatility makes sense.


### 2.2 Write a function that computes the optimal weights ω∗t+1 based on the inputs μ,Σ,γ,λ and the current weights ωt+. You can assume γ = 4 throughout the entire assignment.

```{r}
# Function to compute optimal portfolio weigths based on inputs. 

efficient_weights <- function(mu, sigma, gamma = 4, 
                              lambda, w_prev
) {
  
  iota <- rep(1, ncol(sigma))
  
  sigma_adjusted <- sigma + lambda/gamma * diag(ncol(sigma))
  mu_adjusted <- mu + lambda*w_prev
  
  sigma_inv <- solve(sigma_adjusted)
  
  w_mvp <- sigma_inv %*% iota
  w_mvp <- as.vector(w_mvp / sum(w_mvp))
  
  w_opt <- w_mvp + 1/gamma*(sigma_inv - (sigma_inv %*% iota %*% t(iota) %*% sigma_inv)/sum(sigma_inv)) %*% mu_adjusted
  return(as.vector(w_opt))
}

```


#### 2.3 Use the full sample of your data to compute sample mean returns and the sample variance covariance matrix.

```{r}
# We will need a a (T x N) matrix of returns to compute a matrix of returns: 
ret_matrix <- data %>%
  group_by(permno) %>%
  select(-industry) %>%
  pivot_wider(
    names_from = permno,
    values_from = ret_excess
  ) %>% 
  select(-month) %>%
  drop_na() 


# using the returns matrix, we can compute the covariance matrix 
sigma <- cov(ret_matrix)
mu <- colMeans(ret_matrix) # extract the sample mean returns from the return matrix

remove(data)

```

#### 2.3 Suppose that the initial portfolio ωt+is the naive, equal weighted allocation and the investor has a risk aversion of γ = 4. Compute the optimal portfolio ω∗t+1(λ) for different values of the transaction cost parameter λ.
```{r, fig.width=8, fig.height=4}
# To calculate the optimal portfolio with 1/N initial we will use the hard-coded weight parameter from the function, as this provides an initially equal-weighted portfolio.

naive_weights = 1/ncol(sigma) * rep(1,ncol(sigma))

transactions_costs <- expand_grid(lambda = 20 * qexp((1:99)/100)) %>%
  mutate(weights = map(.x = lambda,
                       ~efficient_weights(
                         sigma = sigma,
                         mu = mu,
                         lambda = .x / 10000,
                         w_prev = naive_weights
                         
                       )
                       ),
         concentration = map_dbl(weights, ~sum(abs(. - 1/ncol(sigma) * rep(1,ncol(sigma)))))
         )

transactions_costs %>% 
  ggplot(aes(x = lambda, y = concentration)) + 
  geom_line(color="tomato3") +
  labs(x = "Transaction cost parameter", 
       y = "Distance from naive",
       title = "Optimal portfolio weights for different transaction cost")

remove(transactions_costs, mu, sigma)

```

The higher the level of transaction costs, the closer will the optimal portfolio be to the equal-weighted. 

### Implement a full-fledged portfolio backtesting strategy with transaction costs proportional to risk as above and suppose that λ = 200/10000 = 200bp. 

```{r}
# For the backtest, we recompute optimal weights just based on past available data.

window_length <- 50
periods <-  nrow(ret_matrix) - window_length
lambda = 200/10000

performance_values <- matrix(NA,
                             nrow = periods,
                             ncol = 3)

colnames(performance_values) = c("raw_return","turnover","net_return")


performance_values <- list("MV (TC)" = performance_values, 
                           "MV (No SS)" = performance_values, 
                            "Naive" = performance_values)

# Helper function to adjust weights due to returns changing
adjust_weights <- function(w, next_return){
  w_prev <- 1 + w * next_return
  as.numeric(w_prev / sum(as.vector(w_prev)))
}

# and helper function to calculate performance evaluation: compute realized returns net of transaction costs.
evaluate_performance <- function(w, w_previous, next_return, lambda){
  raw_return <- as.matrix(next_return) %*% w
  turnover <- sum(abs(w - w_previous))
  net_return <- raw_return - lambda * turnover
  c(raw_return, turnover, net_return)
}

```


```{r}
# The following code chunk performs rolling-window estimation:

for (p in 1:periods) {
  
  returns_window <- ret_matrix[p : (p + window_length - 1), ] # window of return rows. window of rows and all columns
  one_ahead_return <- ret_matrix[p + window_length,] # Next periods return
  
  sigma <- cov(returns_window)
  mu  <- colMeans(returns_window)
  
  # Transaction cost-adjusted portfolio:
  w_MV_TC <- efficient_weights(mu = mu,
                                    sigma = sigma,
                                    lambda = lambda,
                                    w_prev = naive_weights
                                    )
  
  performance_values[[1]][p,] <- evaluate_performance(w = w_MV_TC,
                                                      w_previous = naive_weights,
                                                      next_return = one_ahead_return,
                                                      lambda = lambda
                                                      )
  
  naive_weights = adjust_weights(w_MV_TC, one_ahead_return) # In the first period, this is the naive weight (1/N), but in the second and forward we adjust it. 
  
  
}


```


